True versus apparent shapes of bow shocks

### 2018-03-06 papers/Papers1234/quadrics-bowshock.tex
Submitted first paper! To MNRAS and arxiv
Add new appendixes to address referee points
Explain about the plane of sky as a projective space
Finish the new section on the observations

@@

\newcommand\AddressCRyA{Instituto de Radioastronom\'{\i}a y Astrof\'{\i}sica,
  Universidad Nacional Aut\'onoma de M\'exico, Apartado Postal 3-72,
  58090 Morelia, Michoac\'an, M\'exico}
\author[Tarango Yong \& Henney]{
  Jorge A. Tarango Yong\thanks{E-mail: j.tarango@irya.unam.mx}
  \& William J. Henney\thanks{E-mail: w.henney@irya.unam.mx}\\
  \AddressCRyA
}

@@

\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\Sin}{\mathcal{S}}
\DeclareMathOperator{\Cos}{\mathcal{C}}
\DeclareMathOperator{\Cot}{\mathcal{T}}
\DeclareMathOperator{\GammaFunc}{\Gamma}
\DeclareMathOperator{\Depart}{\Delta}
\newcommand\ecc{\ensuremath{e}}
\newcommand\w{\ensuremath{\mathrm{w}}}
\newcommand\C{\ensuremath{\mathrm{c}}}
\providecommand{\abs}[1]{\lvert#1\rvert}
\providecommand{\Abs}[1]{\left\lvert#1\right\rvert}
\newcommand\TODO[1]{%
  \begin{center}
    \framebox{\parbox{0.8\linewidth}{
        \texttt{\footnotesize\color{red} #1}}}
  \end{center}}

\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle
\begin{abstract}
  Astrophysical bow shocks are a common result of the interaction
  between two supersonic plasma flows, such as winds or jets from
  stars or active galaxies, or streams due to the relative motion
  between a star and the interstellar medium.
  % Similarly shaped radiation bow waves are produced
  % by the relative motion between dust grains and a luminous source.
  For cylindrically symmetric bow shocks, we develop a general theory
  for the effects of inclination angle on the apparent shape. We
  propose a new two-dimensional classification scheme for bow shapes,
  which is based on dimensionless geometric ratios that can be
  estimated from observational images.  The two ratios are related to
  the flatness of the bow's apex, which we term \textit{planitude} and
  the openness of its wings, which we term \textit{alatude}.  We
  calculate the expected distribution in the planitude--alatude plane
  for a variety of simple geometrical and physical models: quadrics of
  revolution, wilkinoids, cantoids, and ancantoids.  We further test
  our methods against numerical magnetohydrodynamical simulations of
  stellar bow shocks and find that the apparent planitude and alatude
  measured from infrared dust continuum maps serve as accurate
  diagnostics of the shape of the contact discontinuity, which can be
  used to discriminate between different physical models.
\end{abstract}

@@

We are grateful for financial support provided by Dirección General de
Asuntos del Personal Académico, Universidad Nacional Autónoma de
México, through grant Programa de Apoyo a Proyectos de Investigación e
Inovación Tecnológica IN111215.  JATY acknowledges support via a
research studentship from Consejo Nacional de Ciencias y Tecnología,
Mexico.

\bibliographystyle{mnras}
%% All references should be put in the BibTeX file bowshocks-biblio.bib
\bibliography{bowshocks-biblio}
\appendix
\input{app-parabola}
%\input{app-ancantoid}
\input{app-shape-parameters}
% \input{app-dust-equations}
\input{app-rcurv-empirical}
% Don't change these lines
\bsp	% typesetting comment
\label{lastpage}
\end{document}

>>>>

% Alternative hollow blackboard bold fonts
\usepackage{dsfont}
\usepackage{bbm}

@@

\newcommand\AddressCRyA{Instituto de Radioastronom\'{\i}a y Astrof\'{\i}sica,
  Universidad Nacional Aut\'onoma de M\'exico, Apartado Postal 3-72,
  58090 Morelia, Michoac\'an, M\'exico}
\author[Tarango-Yong \& Henney]{
  Jorge A. Tarango-Yong\thanks{E-mail: j.tarango@irya.unam.mx}
  \& William J. Henney\thanks{E-mail: w.henney@irya.unam.mx}\\
  \AddressCRyA
}

@@

\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\Sin}{\mathcal{S}}
\DeclareMathOperator{\Cos}{\mathcal{C}}
\DeclareMathOperator{\Cot}{\mathcal{T}}
\DeclareMathOperator{\GammaFunc}{\Gamma}
\DeclareMathOperator{\Depart}{\Delta}
\newcommand\ecc{\ensuremath{e}}
\newcommand\w{\ensuremath{\mathrm{w}}}
\newcommand\C{\ensuremath{\mathrm{c}}}
\providecommand{\abs}[1]{\lvert#1\rvert}
\providecommand{\Abs}[1]{\left\lvert#1\right\rvert}
\newenvironment{Vector}{\left(\begin{array}{c}}{\end{array}\right)}
\newcommand\uvec[1]{\bm{\hat{#1}}}
\newcommand\T{_{\mathrm{\scriptscriptstyle T}}}

\newcommand\TODO[1]{%
  \begin{center}
    \framebox{\parbox{0.8\linewidth}{
        \texttt{\footnotesize\color{red} #1}}}
  \end{center}}

\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle
\begin{abstract}
  Astrophysical bow shocks are a common result of the interaction
  between two supersonic plasma flows, such as winds or jets from
  stars or active galaxies, or streams due to the relative motion
  between a star and the interstellar medium.
  % Similarly shaped radiation bow waves are produced
  % by the relative motion between dust grains and a luminous source.
  For cylindrically symmetric bow shocks, we develop a general theory
  for the effects of inclination angle on the apparent shape. We
  propose a new two-dimensional classification scheme for bow shapes,
  which is based on dimensionless geometric ratios that can be
  estimated from observational images.  The two ratios are related to
  the flatness of the bow's apex, which we term \textit{planitude} and
  the openness of its wings, which we term \textit{alatude}.  We
  calculate the expected distribution in the planitude--alatude plane
  for a variety of simple geometrical and physical models: quadrics of
  revolution, wilkinoids, cantoids, and ancantoids.  We further test
  our methods against numerical magnetohydrodynamical simulations of
  stellar bow shocks and find that the apparent planitude and alatude
  measured from infrared dust continuum maps serve as accurate
  diagnostics of the shape of the contact discontinuity, which can be
  used to discriminate between different physical models.  We present
  an algorithm that can determine the planitude and alatude from observed
  bow shock emission maps with a precision of 10 to 20\%. 
\end{abstract}

@@

We are grateful for financial support provided by Dirección General de
Asuntos del Personal Académico, Universidad Nacional Autónoma de
México, through grant Programa de Apoyo a Proyectos de Investigación e
Inovación Tecnológica IN111215.  JATY acknowledges support via a
research studentship from Consejo Nacional de Ciencias y Tecnología,
Mexico.  This work has made extensive use of Python language libraries
from the SciPy \citep{Jones:2001a} and AstroPy
\citep{Astropy-Collaboration:2013a, Astropy-Collaboration:2018a}
projects.

\bibliographystyle{mnras}
%% All references should be put in the BibTeX file bowshocks-biblio.bib
\bibliography{bowshocks-biblio}
\appendix
\input{app-intermediate}
\input{app-parabola}
%\input{app-ancantoid}
\input{app-shape-parameters}
% \input{app-dust-equations}
\input{app-rcurv-empirical}
% Don't change these lines
\bsp	% typesetting comment
\label{lastpage}
\end{document}

### 2018-03-06 papers/Papers1234/sec-conclusions.tex
Finish the new section on the observations

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figs/new-two-arc-multi-fig}
  \caption[]{Example empirical determination of planitude and alatude
    for observed bow shocks associated with proplyds in the Orion
    Nebula (M42).  Color scale and contours show an \textit{HST}
    H\(\alpha\) image (ACS F658N filter, \citealp{Bally:2006a}) of
    M42~000-400 (panels \textit{a--c}) and M42~069-601 (panels
    \textit{d--f}).  The image scale and orientation are indicated on
    panel~\textit{a} and are the same for all panels.  Three different
    bows have been traced by eye on each object (red and white filled
    symbols): (\textit{a, d})~inner edge, (\textit{b, e})~ridge of
    maximum emission, and (\textit{c, f})~outer edge.  For each panel,
    the dark-colored circle shows the initial fit to the full set of
    points (white and red), using the algorithm described in
    Appendix~\ref{app:rcurv-empirical}.  The center of curvature and
    derived axis are shown by a small filled circle and dashed line in
    the same color. Lighter colored circles show three subsequent
    iterations where the fit is restricted to points within
    \(\pm \Delta\theta = \ang{75}\) of the axis.  The subset of points used in
    the final iteration is marked in red.  The perpendicular radii for
    the final iteration are shown by dotted lines. In panels
    \textit{a, b, d--f} the iterations converge immediately, but in
    panel \textit{c} the iterations stably oscillate between two
    slightly different solutions. }
  \label{fig:000-400-fit}
\end{figure*}

@@

As an example of measuring the projected planitude and alatude of real
bow shocks, we present an analysis of M42~000-400 and M42~069-601,
which are two H\(\alpha\) emission arcs \citep{Bally:2000a,
  Gutierrez-Soto:2015a} associated with proplyds\footnote{%
  The coordinate designation (see \citealp{ODell:1994a} for an
  explanation of the nomenclature) of 000-400 is very imprecise in
  right ascension, but we use it here for consistency with previous
  papers. The associated proplyd is listed with the more correct
  designation 4596-400 in catalogs such as \citet{Ricci:2008a}.  } %
in the west of the Orion Nebula (M42).  An image of one of these arcs
(M42~069-601) was used in the illustration of bow shock terminology in
Figure~\ref{fig:bow-terminology}.

@@

We determine the planitude and alatude by fitting a circle to the
traced points within \(\pm \Delta\theta = \ang{75}\) of the bow axis, using the
iterative algorithm described in Appendix~\ref{app:rcurv-empirical}.
The fitted circle, when combined with the position of the central
source, yields the orientation of the bow axis, together with the apex
distance, \(R_0\), radius of curvature, \(R_\C\), and two
perpendicular radii (one for each wing), \(R_{90+}\) and \(R_{90-}\).
These are all indicated on the panels of Figure~\ref{fig:000-400-fit}
by light-colored lines.\footnote{%
  For conciseness, we drop the prime symbol from the radii, both in
  this section and in Appendix~\ref{app:rcurv-empirical}, even though
  they are all projected quantities.} %
The projected planitude and alatude then follow as
\(\Pi' = R_\C/R_0\), \(\Lambda_+ = R_{90+}/R_0\),
\(\Lambda_- = R_{90-}/R_0\), which are shown in
Figure~\ref{fig:000-400-planitude-alatude}.

@@

It could be argued that much of the between-tracing variations in
\(\Pi'\) and \(\langle\Lambda'\rangle\) are due to real differences in the shape of the
inner and outer boundaries of the emission arc.  Although this may be
true, in the absence of a robust theory as to exactly what feature of
the observed images constitutes \emph{the} bow shock, such variations
nevertheless serve to limit the precision with which the bow shock
shape can be measured.  We therefore conclude that conservative
estimates of \(20\%\) precision for \(\Pi'\) and \(10\%\) precision for
\(\Lambda'\) are appropriate when analyzing observations of a similar or
better quality\footnote{%
  Since the uncertainties are systematic and due to unavoidably
  subjective decisions, it is unlikely that better quality
  observations would improve the situation, although poorer quality
  observations could make things worse.  } %
to those presented in Figure~\ref{fig:000-400-fit}.  This will be an
important limitation when comparing the statistics of the shapes of
different bow shock populations, as we will present in a following
paper.

\subsection{Derived shape of M42 arcs}
\label{sec:derived-shape-m42}

From a casual inspection of Figure~\ref{fig:000-400-fit}, it is
apparent that the shapes of the two M42 arcs are closely similar, and
this is confirmed by the numbers in Table~\ref{tab:m42-fit}. Both
000-400 and 069-601 are consistent with
\((\Pi', \Lambda') \approx (3.0, 2.5)\), and if we take the absolute minimum over
all the different tracings and reasonable variations in the
\(\Delta\theta\) fit parameter, we find firm lower limits of
\((\Pi', \Lambda') > (2.1, 2.1)\).  In the rest of this section

If we compare th

The principal hindrance to an accurate tracing of the bow shape in
this object is the strong and variable nebular background emission,
which particularly affects the far wing on the right hand side of the
image, where a bright background filament is superimposed on the bow.

These give an idea of the systematic uncertainties and sensitivity to
subjective choices,

@@

Finally (\S~\ref{sec:more-realistic-bow}), we test the projected shape
analysis methods against the results of computational fluid dynamic
simulations of magnetized and non-magnetized bow shocks from
\citet{Meyer:2017a} of a runaway OB main-sequence star.  We find that
measurements made on maps of infrared dust emission can be accurate
diagnostics of the projected shape of the contact discontinuity for
this type of bow shock (Fig.~\ref{fig:sim-Pi-Lambda}).  The
distributions of projected planitude and alatude for a population of
randomly oriented bow shocks shows systematic differences between the
different simulations.

>>>>

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figs/new-two-arc-multi-fig}
  \caption[]{Example empirical determination of planitude and alatude
    for observed bow shocks associated with proplyds in the outer
    Orion Nebula (M42).  Color scale and contours show an \textit{HST}
    H\(\alpha\) image (ACS F658N filter, \citealp{Bally:2006a}) of
    M42~000-400 (panels \textit{a--c}) and M42~069-601 (panels
    \textit{d--f}).  The image scale and orientation are indicated on
    panel~\textit{a} and are the same for all panels.  Three different
    bows have been traced by eye on each object (red and white filled
    symbols): (\textit{a, d})~inner edge, (\textit{b, e})~ridge of
    maximum emission, and (\textit{c, f})~outer edge.  For each panel,
    the dark-colored circle shows the initial fit to the full set of
    points (white and red), using the algorithm described in
    Appendix~\ref{app:rcurv-empirical}.  The center of curvature and
    derived axis are shown by a small filled circle and dashed line in
    the same color. Lighter colored circles show three subsequent
    iterations where the fit is restricted to points within
    \(\pm \Delta\theta = \ang{75}\) of the axis.  The subset of points used in
    the final iteration is marked in red.  The perpendicular radii for
    the final iteration are shown by dotted lines. In panels
    \textit{a, b, d--f} the iterations converge immediately, but in
    panel \textit{c} the iterations stably oscillate between two
    slightly different solutions. }
  \label{fig:000-400-fit}
\end{figure*}

@@

As an example of measuring the projected planitude and alatude of real
bow shocks, we present an analysis of M42~000-400 and M42~069-601,
which are two H\(\alpha\) emission arcs \citep{Bally:2000a,
  Gutierrez-Soto:2015a} associated with proplyds\footnote{%
  The coordinate designation (see \citealp{ODell:1994a} for an
  explanation of the nomenclature) of 000-400 is very imprecise in
  right ascension, but we use it here for consistency with previous
  papers. The associated proplyd is listed with the more correct
  designation 4596-400 in catalogs such as \citet{Ricci:2008a}.  } %
in the west of the Orion Nebula (M42) at a distance of roughly
\SI{0.5}{pc} from the high-mass Trapezium stars that ionize the
nebula.  An image of one of these arcs (M42~069-601) was used in the
illustration of bow shock terminology in
Figure~\ref{fig:bow-terminology}.

@@

We determine the planitude and alatude by fitting a circle to the
traced points within \(\pm \Delta\theta = \ang{75}\) of the bow axis, using the
iterative algorithm described in Appendix~\ref{app:rcurv-empirical}.
The fitted circle, when combined with the position of the central
source, yields the orientation of the bow axis, together with the apex
distance, \(R_0\), radius of curvature, \(R_\C\), and two
perpendicular radii (one for each wing), \(R_{90+}\) and \(R_{90-}\).
These are all indicated on the panels of Figure~\ref{fig:000-400-fit}
by light-colored lines.\footnote{%
  For conciseness, we drop the prime symbol from the radii, both in
  this section and in Appendix~\ref{app:rcurv-empirical}, even though
  they are all projected quantities.} %
The projected planitude and alatude then follow as
\(\Pi' = R_\C/R_0\), \(\Lambda_+' = R_{90+}/R_0\),
\(\Lambda_-' = R_{90-}/R_0\), which are shown in
Figure~\ref{fig:000-400-planitude-alatude}.

@@

It could be argued that much of the between-tracing variations in
\(\Pi'\) and \(\langle\Lambda'\rangle\) are due to real differences between the shapes of
the inner and outer boundaries of the emission arc.  Although this may
be true, in the absence of a robust theory as to exactly what feature
of the observed images constitutes \emph{the} bow shock, such
variations nevertheless serve to limit the precision with which the
bow shock shape can be measured.  We therefore conclude that
conservative estimates of \(20\%\) precision for \(\Pi'\) and \(10\%\)
precision for \(\Lambda'\) are appropriate when analyzing observations of a
similar or better quality\footnote{%
  Since the uncertainties are systematic and due to unavoidably
  subjective decisions, it is unlikely that better quality
  observations would improve the situation, although poorer quality
  observations could make things worse.  } %
to those presented in Figure~\ref{fig:000-400-fit}.  This will be an
important limitation when comparing the statistics of the shapes of
different bow shock populations, as we will present in a following
paper.

\subsection{Derived shape of the M42 arcs}
\label{sec:derived-shape-m42}

From a casual inspection of Figure~\ref{fig:000-400-fit}, it is
apparent that the shapes of the two M42 arcs are closely similar, and
this is confirmed by the numbers in Table~\ref{tab:m42-fit}. Both
000-400 and 069-601 are consistent with
\((\Pi', \Lambda') \approx (3.0, 2.5)\), and if we take the absolute minimum over
all the different tracings and reasonable variations in the
\(\Delta\theta\) fit parameter, we find unassailable lower limits of
\((\Pi', \Lambda') > (2.1, 2.1)\).  In the rest of this section, we consider
only these lower limits, since they are already sufficient for drawing
interesting conclusions.

Comparison with Figure~\ref{fig:sim-Pi-Lambda} shows that both the MHD
and HD simulations of \citet{Meyer:2016a} are inconsistent with the
observations.  Although \(\Pi' > 2.1\) can be satisfied,\footnote{%
  Either by the MHD simulations at low inclinations
  (\(\abs{i} = \ang{0}\)--\ang{20}) or by the HD simulation at
  intermediate inclinations (\(\abs{i} = \ang{30}\)--\ang{40}).} %
the simulations' projected alatude is \(\Lambda' < 1.9\) for all
inclinations, which is significantly less than the observed lower
limit of \(2.1\).  This is not particularly surprising since the
simulations were not tailored to the situation of these proplyd bow
shocks in M42, in which the mildly supersonic photoevaporation flow
from an externally irradiated protoplanetary disk interacts with the
mildly supersonic champagne flow from the core of the Orion Nebula.
The proplyd case has at least four important differences from the
runaway O-star case modeled by \citet{Meyer:2016a}: (1)~The velocity
of the outer wind is \(\le \SI{20}{km.s^{-1}}\) instead of
\SI{40}{km.s^{-1}}; (2)~The outer wind is slightly divergent, rather
than plane-parallel; (3)~Both inner and outer shocks are strongly
radiative, so both shells (see Fig.~\ref{fig:2-winds}) contribute to
the observed emission; (4)~The inner wind is not isotropic, but
instead corresponds to the \(k = 0.5\) case of
equation~\eqref{eq:ancantoid-density}.  The ways in which these
differences may account for the discrepancy with the observations will
be explored in detail in a subsequent paper.

@@

In the following section (\S~\ref{sec:more-realistic-bow}), we test the
projected shape analysis methods against the results of computational
fluid dynamic simulations of magnetized and non-magnetized bow shocks
from \citet{Meyer:2017a} of a runaway OB main-sequence star.  We find
that measurements made on maps of infrared dust emission can be
accurate diagnostics of the projected shape of the contact
discontinuity for this type of bow shock
(Fig.~\ref{fig:sim-Pi-Lambda}).  The distributions of projected
planitude and alatude for a population of randomly oriented bow shocks
shows systematic differences between the different simulations.

Finally (\S~\ref{sec:obs}), we give an example of the application of
our methods to observed emission maps of bow shocks, describing a
robust algorithm for empirically determining the projected planitude
and alatude from imperfect real data.  We investigate the sensitivity
of the results to systematic errors due to both observational
uncertainties and subjective choices in the application of the
algorithm.  We find that the projected planitude and alatude can be
determined with precisions of 20\% and 10\%, respectively.  For our
illustrative observations, we show that this is more than sufficient
to rule out certain models.
